Lesson 5
========================================================

### Multivariate Data
Notes:

***

### Moira Perceived Audience Size Colored by Age
Notes:

***

### Third Qualitative Variable
Notes:

```{r Third Qualitative Variable}
library(ggplot2) 
library(ggthemes) 
theme_set(theme_minimal(24)) 
pf <- read.csv( "pseudo_facebook.tsv", sep = '\t')
ggplot(aes(x = gender, y = age),
       data = subset(pf, !is.na(gender))) + geom_boxplot() + 
  stat_summary(fun.y = mean, geom = "point", shape = 4)

ggplot(aes(x = gender, y = age),
       data = subset(pf, !is.na(gender))) + geom_histogram()

ggplot(aes(x = age, y = friend_count),
       data = subset(pf, !is.na(gender))) + 
  geom_line(aes(color = gender), stat = "summary",
            fun.y = median)


```

# Write code to create a new data frame,
# called 'pf.fc_by_age_gender', that contains
# information on each age AND gender group.

# The data frame should contain the following variables:

#    mean_friend_count,
#    median_friend_count,
#    n (the number of users in each age and gender grouping)

# Here is an example of the structure of your data frame. Your
# data values will be different. Note that if you are grouping by
# more than one variable, you will probably need to call the
# ungroup() function. 

#   age gender mean_friend_count median_friend_count    n
# 1  13 female          247.2953                 150  207
# 2  13   male          184.2342                  61  265
# 3  14 female          329.1938                 245  834
# 4  14   male          157.1204                  88 1201

# See the Instructor Note for two hints.

# DO NOT DELETE THESE NEXT TWO LINES OF CODE
# ==============================================================
pf <- read.delim('/datasets/ud651/pseudo_facebook.tsv')
suppressMessages(library(dplyr))

# ENTER YOUR CODE BELOW THIS LINE.
# ==============================================================
```{r}
library(dplyr)
pf.fc_by_age_gender <- pf %>%
  filter(!is.na(gender)) %>%
  group_by(age, gender) %>%
  summarise( mean_friend_count = mean(friend_count), 
             median_friend_count = median(as.numeric(friend_count)), 
             n = n()) %>%
  ungroup() %>%
  arrange(age)

```


***

### Plotting Conditional Summaries
Notes:
# Create a line graph showing the
# median friend count over the ages
# for each gender. Be sure to use
# the data frame you just created,
# pf.fc_by_age_gender.
```{r Plotting Conditional Summaries}

ggplot(aes(x = age, y = median_friend_count), 
              data = subset(pf.fc_by_age_gender, !is.na(gender))) + 
  geom_line(aes(color = gender))
```

***

### Thinking in Ratios
Notes:

***

### Wide and Long Format
Notes:

```{r}
install.packages("tidyr") # only need to run this once 
library(tidyr) 
spread(subset(pf.fc_by_age_gender[c('age', 'gender', 'median_friend_count')], 
              !is.na(gender)), gender, median_friend_count)
```


***

### Reshaping Data
Notes:

```{r}
install.packages('reshape2')
library(reshape2)

pf.fc_by_age_gender.wide <- dcast(pf.fc_by_age_gender,
                                  age ~ gender,
                                  value.var = 'median_friend_count')

melt(pf.fc_by_age_gender.wide, id.vars = "age", 
     measure.vars = c("male", "female"), 
     value.name = "median_friend_count",
     variable.name = 'gender')
```


***

### Ratio Plot
Notes:

# Plot the ratio of the female to male median
# friend counts using the data frame
# pf.fc_by_age_gender.wide.

# Think about what geom you should use.
# Add a horizontal line to the plot with
# a y intercept of 1, which will be the
# base line. Look up the documentation
# for geom_hline to do that. Use the parameter
# linetype in geom_hline to make the
# line dashed.

# The linetype parameter can take the values 0-6:
# 0 = blank, 1 = solid, 2 = dashed
# 3 = dotted, 4 = dotdash, 5 = longdash
# 6 = twodash

# This assignment is not graded and
# will be marked as correct when you submit.

# ENTER YOUR CODE BELOW THIS LINE
# =================================================

```{r Ratio Plot}
ggplot(aes(x = age, y = female/male), 
              data = pf.fc_by_age_gender.wide) + 
  geom_line() + 
  geom_hline(yintercept=1, alpha = 0.3, linetype = 2)


```

***

### Third Quantitative Variable
Notes:

# Create a variable called year_joined
# in the pf data frame using the variable
# tenure and 2014 as the reference year.

# The variable year joined should contain the year
# that a user joined facebook.

# See the Instructor Notes for three hints if you get
# stuck. Scroll down slowly to see one hint at a time
# if you would like some guidance.

# This programming exercise WILL BE automatically graded.

# DO NOT ALTER THE CODE BELOW THIS LINE
# ========================================================
pf <- read.delim('/datasets/ud651/pseudo_facebook.tsv')

# ENTER YOUR CODE BELOW THIS LINE.
# ========================================================


```{r Third Quantitative Variable}

pf$year_joined <- floor(2014 - pf$tenure/365)

```

***

### Cut a Variable
Notes:

# Create a new variable in the data frame
# called year_joined.bucket by using
# the cut function on the variable year_joined.

# You need to create the following buckets for the
# new variable, year_joined.bucket

#        (2004, 2009]
#        (2009, 2011]
#        (2011, 2012]
#        (2012, 2014]

# Note that a parenthesis means exclude the year and a
# bracket means include the year.

# Look up the documentation for cut or try the link
# in the Instructor Notes to accomplish this task.

```{r Cut a Variable}
summary(pf$year_joined)
table(pf$year_joined)

pf$year_joined.bucket <- cut(pf$year_joined, 
                             breaks = c(2004, 2009, 2011, 2012, 2014))
```

***

### Plotting it All Together
Notes:

# Create a line graph of friend_count vs. age
# so that each year_joined.bucket is a line
# tracking the median user friend_count across
# age. This means you should have four different
# lines on your plot.

# You should subset the data to exclude the users
# whose year_joined.bucket is NA.

# If you need a hint, see the Instructor Notes.

# This assignment is not graded and
# will be marked as correct when you submit.

```{r Plotting it All Together}
ggplot(aes(x = age, y = friend_count), 
              data = subset(pf, !is.na(year_joined))) + 
  geom_line(aes(color = year_joined.bucket), stat = 'summary', fun.y = median)
```

***

### Plot the Grand Mean
Notes:

# Write code to do the following:

# (1) Add another geom_line to code below
# to plot the grand mean of the friend count vs age.

# (2) Exclude any users whose year_joined.bucket is NA.

# (3) Use a different line type for the grand mean.

# As a reminder, the parameter linetype can take the values 0-6:

# 0 = blank, 1 = solid, 2 = dashed
# 3 = dotted, 4 = dotdash, 5 = longdash
# 6 = twodash

# This assignment is not graded and
# will be marked as correct when you submit.

# The code from the last programming exercise should
# be your starter code!


```{r Plot the Grand Mean}
ggplot(aes(x = age, y = friend_count), 
              data = subset(pf, !is.na(year_joined))) + 
  geom_line(aes(color = year_joined.bucket), 
            stat = 'summary', 
            fun.y = mean) +
  geom_line(stat = 'summary', 
            fun.y = mean, 
            linetype = 2)
```

***

### Friending Rate
Notes:
What is the median friend rate?
0.2205 
What is the maximum friend rate?
417.0000
```{r Friending Rate}
summary(with(subset(pf, tenure > 0), friend_count/tenure))
```

***

### Friendships Initiated
Notes:

# Create a line graph of mean of friendships_initiated per day (of tenure)
# vs. tenure colored by year_joined.bucket.

# You need to make use of the variables tenure,
# friendships_initiated, and year_joined.bucket.

# You also need to subset the data to only consider user with at least
# one day of tenure.

```{r Friendships Initiated}
ggplot(aes(x = tenure, y = friendships_initiated/tenure), 
              data = subset(pf, tenure > 0)) + 
  geom_line(aes(color = year_joined.bucket), 
            stat = 'summary', 
            fun.y = mean)
```

***

### Bias-Variance Tradeoff Revisited
Notes:

# Instead of geom_line(), use geom_smooth() to add a smoother to the plot.
# You can use the defaults for geom_smooth() but do color the line
# by year_joined.bucket

```{r Bias-Variance Tradeoff Revisited}

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_smooth(aes(color = year_joined.bucket))

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

```

***

### Sean's NFL Fan Sentiment Study
Notes:

***

### Introducing the Yogurt Data Set
Notes:

***

### Histograms Revisited
Notes:

```{r Histograms Revisited}

yo <- read.csv('yogurt.csv')
str(yo)

# Change the id from an int to a factor
yo$id <- factor(yo$id)
str(yo)

ggplot(aes(x = price), data = yo) +
  geom_histogram(binwidth = 10)

```

***

### Number of Purchases
Notes:

# Create a new variable called all.purchases,
# which gives the total counts of yogurt for
# each observation or household.

# One way to do this is using the transform
# function. You can look up the function transform
# and run the examples of code at the bottom of the
# documentation to figure out what it does.

# The transform function produces a data frame
# so if you use it then save the result to 'yo'!

# OR you can figure out another way to create the
# variable.

```{r Number of Purchases}
summary(yo)
length(unique(yo$price))
table(yo$price)

yo <- transform(yo, all.purchases = strawberry + blueberry + pina.colada + plain + mixed.berry)

```

***

### Prices over Time
Notes:

# Create a scatterplot of price vs time.

# This will be an example of a time series plot.

# Resolve overplotting issues by using
# techniques you learned in Lesson 4.

# What are some things that you notice?

```{r Prices over Time}

ggplot(aes(x = time, y = price), data = yo) +
  geom_point(alpha = 1/10, position = position_jitter ())
# or
ggplot(aes(x = time, y = price), data = yo) +
  geom_jitter(alpha = 1/10, shape = 21, fill = 'orange')
```

***

### Sampling Observations
Notes:

***

### Looking at Samples of Households

```{r Looking at Sample of Households}
set.seed(4232)
sample.ids <- sample(levels(yo$id), 16)

ggplot(aes(x = time, y = price), 
       data = subset(yo, id %in% sample.ids)) +
  facet_wrap( ~ id) + 
  geom_line() +
  geom_point(aes(size = all.purchases), pch = 1)
```

There's a lot of variation among households in terms of yogurt purchases. Without coupon or sale data it's hard to gain many insights. In general, most of the large purchaes occur below the 50 cent mark, although not exclusively. This could indicate that some people are more likely to stock up when the prices are low, while others are more consistent in their purchases, regardless of price. It likely comes down to how important yogurt is them to have in their diet. 

***

### The Limits of Cross Sectional Data
Notes:

***

### Many Variables
Notes:

***

### Scatterplot Matrix
Notes:

```{r}

install.packages('GGally')
library(GGally)
theme_set(theme_minimal(20))

set.seed(1836)

pf_subset <- pf[ , c(2:15)] 
names(pf_subset)
ggpairs(pf_subset[sample.int(nrow(pf_subset, 1000))], axisLabels = 'internal')

```


***

### Even More Variables
Notes:

***

### Heat Maps
Notes:

```{r}
nci <- read.table("nci.tsv")
colnames(nci) <- c(1:64)
```

```{r}
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))
```


***

### Analyzing Three of More Variables
Reflection:

I learned how to explore the relationships between more than two variables by creating new variables, slicing and dicing data to gain new insights, dealing with noise vs bias tradeoffs, sampling a dataset to explore relationships, and visualizing relationships between multiple variables at once. 

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

